---
title: "Markov Chain Modelling of No-Claims Discount Systems"
author: "Ntando Sithole"
date: "2026-06-20"
description: "A reproducible R framework for projecting rating-class migration, premium income, and the long-run portfolio mix of a no-claims discount system."
image: ../../assets/markov-ncd-cover.svg
image-alt: "Abstract transition paths connecting four no-claims discount classes"
categories:
  - Stochastic Processes
  - Actuarial Modelling
  - General Insurance
  - R
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-copy: true
execute:
  freeze: true
  echo: true
  warning: false
  message: false
---

## Introduction

No-claims discount (NCD) systems adjust a policyholder's renewal premium
according to recent claims experience. A policyholder typically moves toward a
higher-discount class after a claim-free year and toward a lower-discount class
after one or more claims.

A discrete-time Markov chain provides a transparent framework for modelling
these movements. The approach supports several practical actuarial tasks,
including:

- projecting the future distribution of policyholders across rating classes;
- estimating expected premium income over a planning horizon;
- assessing the long-run mix of an NCD portfolio; and
- testing the financial effect of alternative transition rules or discount
  scales.

This article develops an illustrative four-class NCD model using reproducible
base R code. The assumptions are synthetic and are intended to demonstrate the
method rather than represent a calibrated insurance portfolio.

## Methodology

Let \(X_t\) denote the NCD class occupied at the start of policy year \(t\).
The state space is

$$
\mathcal{S} = \{1,2,3,4\},
$$

where Class 1 carries no discount and Class 4 carries the highest discount.
Under the first-order Markov assumption,

$$
\Pr(X_{t+1}=j \mid X_t=i, X_{t-1},\ldots,X_0)
= \Pr(X_{t+1}=j \mid X_t=i)
= p_{ij}.
$$

The one-year transition matrix is

$$
\mathbf{P}
=
\begin{pmatrix}
0.17 & 0.83 & 0.00 & 0.00 \\
0.17 & 0.00 & 0.83 & 0.00 \\
0.05 & 0.12 & 0.00 & 0.83 \\
0.00 & 0.00 & 0.05 & 0.95
\end{pmatrix}.
$$

Each row sums to one. Entry \(p_{ij}\) is the probability that a policyholder
in Class \(i\) at the start of one year occupies Class \(j\) at the start of the
next year.

If \(\boldsymbol{q}_0\) is the initial row vector of class probabilities, the
distribution after \(t\) transitions is

$$
\boldsymbol{q}_t = \boldsymbol{q}_0\mathbf{P}^t.
$$

For a base annual premium \(B\) and discount vector
\(\boldsymbol{d}=(d_1,\ldots,d_4)^\mathsf{T}\), the premium vector is

$$
\boldsymbol{b}
= B(\boldsymbol{1}-\boldsymbol{d}).
$$

The expected premium payable in policy year \(t+1\) is therefore

$$
\operatorname{E}[B_{t+1}]
= \boldsymbol{q}_0\mathbf{P}^{t}\boldsymbol{b},
$$

and expected cumulative premium over \(n\) years is

$$
\operatorname{E}[C_n]
= \sum_{t=0}^{n-1}
  \boldsymbol{q}_0\mathbf{P}^{t}\boldsymbol{b}.
$$

This timing convention assumes that the premium for each year is determined by
the class occupied at the start of that year.

## Implementation in R

### Transition assumptions

```{r}
#| label: define-transition-matrix

states <- paste("Class", 1:4)

P <- matrix(
  c(
    0.17, 0.83, 0.00, 0.00,
    0.17, 0.00, 0.83, 0.00,
    0.05, 0.12, 0.00, 0.83,
    0.00, 0.00, 0.05, 0.95
  ),
  nrow = 4,
  byrow = TRUE,
  dimnames = list(states, states)
)

stopifnot(all(abs(rowSums(P) - 1) < 1e-12))

knitr::kable(
  P,
  digits = 2,
  caption = "Illustrative annual transition matrix"
)
```

The matrix reflects progressive movement toward higher-discount classes for
policyholders with favourable experience, while retaining a mechanism for
movement toward lower-discount classes.

### Multi-year transition probabilities

The following function calculates \(\mathbf{P}^n\), including
\(\mathbf{P}^0=\mathbf{I}\).

```{r}
#| label: matrix-power

matrix_power <- function(P, n) {
  stopifnot(
    nrow(P) == ncol(P),
    length(n) == 1,
    n >= 0,
    n == as.integer(n)
  )

  result <- diag(nrow(P))

  if (n > 0) {
    for (step in seq_len(n)) {
      result <- result %*% P
    }
  }

  dimnames(result) <- dimnames(P)
  result
}

P_5 <- matrix_power(P, 5)

knitr::kable(
  P_5,
  digits = 4,
  caption = "Five-year transition probabilities"
)
```

The first row of \(\mathbf{P}^5\) gives the class distribution after five
transitions for a policyholder who begins in Class 1.

### Stationary distribution

For an ergodic chain, the stationary distribution
\(\boldsymbol{\pi}\) satisfies

$$
\boldsymbol{\pi}\mathbf{P}=\boldsymbol{\pi},
\qquad
\sum_{j=1}^{4}\pi_j=1.
$$

The system can be solved directly using a constrained linear system.

```{r}
#| label: stationary-distribution

stationary_distribution <- function(P) {
  n_states <- nrow(P)

  coefficient_matrix <- rbind(
    t(P) - diag(n_states),
    rep(1, n_states)
  )

  target <- c(rep(0, n_states), 1)
  stationary <- as.vector(qr.solve(coefficient_matrix, target))

  setNames(stationary, colnames(P))
}

stationary <- stationary_distribution(P)

knitr::kable(
  data.frame(
    Class = names(stationary),
    Probability = as.numeric(stationary)
  ),
  digits = 4,
  caption = "Stationary class distribution"
)
```

Under the illustrative assumptions, approximately \(92.7\%\) of the long-run
portfolio occupies Class 4. This concentration is driven by the high
probability of remaining in Class 4 once it has been reached.

### Expected premium projection

Assume a base annual premium of \(750\) monetary units and class discounts of
\(0\%\), \(15\%\), \(25\%\), and \(30\%\).

```{r}
#| label: premium-assumptions

base_premium <- 750
discounts <- c(0.00, 0.15, 0.25, 0.30)
premium_by_class <- base_premium * (1 - discounts)
names(premium_by_class) <- states

knitr::kable(
  data.frame(
    Class = states,
    Discount = sprintf("%.0f%%", 100 * discounts),
    Premium = premium_by_class
  ),
  caption = "Illustrative NCD premium scale"
)
```

```{r}
#| label: premium-projection-function

expected_premium_projection <- function(
  P,
  initial_distribution,
  premium_by_class,
  years
) {
  stopifnot(
    length(initial_distribution) == nrow(P),
    length(premium_by_class) == ncol(P),
    abs(sum(initial_distribution) - 1) < 1e-12,
    years >= 1
  )

  class_distributions <- matrix(
    NA_real_,
    nrow = years,
    ncol = ncol(P),
    dimnames = list(
      paste("Year", seq_len(years)),
      colnames(P)
    )
  )

  annual_premium <- numeric(years)
  current_distribution <- initial_distribution

  for (year in seq_len(years)) {
    class_distributions[year, ] <- current_distribution
    annual_premium[year] <- sum(
      current_distribution * premium_by_class
    )
    current_distribution <- current_distribution %*% P
  }

  list(
    class_distributions = class_distributions,
    annual_premium = annual_premium,
    cumulative_premium = cumsum(annual_premium)
  )
}

new_policyholder <- setNames(c(1, 0, 0, 0), states)

projection <- expected_premium_projection(
  P = P,
  initial_distribution = new_policyholder,
  premium_by_class = premium_by_class,
  years = 10
)

projection_table <- data.frame(
  Year = seq_len(10),
  Expected_Annual_Premium = projection$annual_premium,
  Expected_Cumulative_Premium = projection$cumulative_premium
)

knitr::kable(
  projection_table,
  digits = 2,
  caption = "Ten-year expected premium projection for a Class 1 policyholder"
)
```

The expected cumulative premium over ten years is approximately \(5{,}825.47\)
monetary units. The annual expected premium declines as probability mass moves
toward the higher-discount classes.

```{r}
#| label: premium-projection-plot
#| fig-cap: "Expected annual premium and its long-run equilibrium level"

long_run_premium <- sum(stationary * premium_by_class)

plot(
  seq_len(10),
  projection$annual_premium,
  type = "b",
  pch = 19,
  xlab = "Policy year",
  ylab = "Expected premium",
  ylim = range(c(projection$annual_premium, long_run_premium))
)

abline(
  h = long_run_premium,
  lty = 2,
  col = "steelblue"
)

legend(
  "topright",
  legend = c("Projected annual premium", "Long-run expected premium"),
  lty = c(1, 2),
  pch = c(19, NA),
  col = c("black", "steelblue"),
  bty = "n"
)
```

The long-run expected annual premium is

$$
\boldsymbol{\pi}\boldsymbol{b}
\approx 529.67.
$$

This is a portfolio-mix result under the model assumptions; it is not, by
itself, a measure of technical profitability.

## Actuarial Interpretation

The model separates two commercially important components:

1. **Policyholder migration:** The transition matrix captures how claims
   experience changes future rating classes.
2. **Financial impact:** The premium vector converts projected class
   distributions into expected income.

This separation makes the framework useful for pricing governance and scenario
analysis. An actuary can change the transition rules without changing the
discount scale, or revise the discount scale while holding policyholder
behaviour constant. The financial effect of each design decision can therefore
be assessed independently.

The stationary distribution also provides a useful diagnostic. A very high
concentration in the maximum-discount class may indicate that the scale is
generous relative to claim frequency, that downward movements are too weak, or
that the model omits lapse and selection effects. In practice, the result
should be compared with observed portfolio distributions and tested under
alternative experience assumptions.

## Model Limitations

The model is intentionally compact. A production implementation would need to
address several limitations:

- **Markov assumption:** Future class movement is assumed to depend only on
  the current class. Claim history beyond the current state may still contain
  predictive information.
- **Time homogeneity:** Transition probabilities are assumed constant.
  Inflation, underwriting actions, economic conditions, and portfolio mix can
  cause them to vary over time.
- **Portfolio heterogeneity:** A single matrix may conceal material differences
  by age, vehicle type, territory, distribution channel, or risk segment.
- **Lapses and new business:** The closed-chain model does not include entry,
  cancellation, non-renewal, or reinstatement.
- **Premium adequacy:** Premium income is projected without claim costs,
  expenses, commission, reinsurance, tax, or the cost of capital.
- **Parameter uncertainty:** Estimated transition probabilities are subject to
  sampling error and should be accompanied by sensitivity or simulation
  analysis.

These limitations do not invalidate the framework. They define the additional
features required before it can support pricing, reserving, planning, or
capital decisions.

## Conclusion

A Markov-chain representation of an NCD system provides a concise link between
policyholder claims experience, rating-class migration, and future premium
income. The model is transparent, auditable, and straightforward to implement
in R.

For practical actuarial use, the next development step would be to estimate
transition matrices from portfolio experience, segment the assumptions by
material risk characteristics, and integrate projected premium with claims,
expenses, lapses, and capital requirements.
