---
title: "Life Insurance Cash Flow Projection and Reserve Sensitivity in R"
author: "Ntando Sithole"
date: "2026-06-24"
description: "A reproducible R model for term assurance cash flows, level premiums, prospective reserves, and sensitivity testing under synthetic mortality assumptions."
image: ../../assets/life-insurance-reserve-cover.svg
image-alt: "Abstract life insurance reserve modelling graphic with a shield, cash-flow bars, and reserve curve"
categories:
  - Life Insurance
  - Actuarial Modelling
  - Valuation
  - IFRS 17
  - R
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-copy: true
execute:
  freeze: true
  echo: true
  warning: false
  message: false
---

## Introduction

Life insurance valuation work is ultimately a disciplined projection exercise:
future premiums, claims, expenses, and reserves are converted into present-value
measures under a clearly stated basis. The modelling can be simple or highly
sophisticated, but the core structure should remain auditable.

This article develops a compact R framework for a level-premium term assurance
portfolio. The objective is not to reproduce a production actuarial system.
Instead, the model demonstrates the mechanics behind:

- mortality-based expected claim projections;
- level premium calculation;
- prospective reserve development by policy duration; and
- sensitivity testing under alternative discount-rate and mortality bases.

The assumptions are synthetic and are included directly in the article so that
the code can be rendered without external data files.

## Methodology

Consider a policy issued to a life aged \(x\), with term \(n\), annual effective
interest rate \(i\), discount factor \(v=(1+i)^{-1}\), and sum assured \(S\).
Let \(q_{x+t}\) denote the one-year mortality rate at attained age \(x+t\).

The probability of surviving from issue to the start of policy year \(t+1\) is

$$
{}_{t}p_x =
\prod_{j=0}^{t-1} (1-q_{x+j}),
\qquad {}_0p_x = 1.
$$

For a benefit payable at the end of the year of death, the actuarial present
value of a term assurance is

$$
A^{1}_{x:\overline{n}|}
=
\sum_{t=0}^{n-1}
v^{t+1} {}_{t}p_x q_{x+t}.
$$

If premiums are payable annually in advance while the policyholder is alive,
the temporary life annuity-due factor is

$$
\ddot{a}_{x:\overline{n}|}
=
\sum_{t=0}^{n-1}
v^t {}_{t}p_x.
$$

The net annual level premium is therefore

$$
P =
\frac{S A^{1}_{x:\overline{n}|}}
{\ddot{a}_{x:\overline{n}|}}.
$$

At duration \(k\), immediately after the \(k\)-th policy anniversary, the
prospective net reserve for an in-force policy is

$$
V_k =
S
\sum_{t=0}^{n-k-1}
v^{t+1} {}_{t}p_{x+k} q_{x+k+t}
-
P
\sum_{t=0}^{n-k-1}
v^t {}_{t}p_{x+k}.
$$

This reserve is a best-estimate technical measure under the selected valuation
basis. In production, it would normally be extended for expenses, lapses,
reinsurance, policyholder options, risk adjustment, contractual service margin,
capital requirements, and governance controls.

## Implementation in R

### Model setup

```{r}
#| label: setup
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

format_currency <- function(x) {
  paste0("R", format(round(x, 0), big.mark = ",", scientific = FALSE))
}

format_percent <- function(x) {
  paste0(round(100 * x, 2), "%")
}

issue_age <- 40
term <- 20
sum_assured <- 1000000
initial_policies <- 1000
annual_interest <- 0.085
annual_expense <- 250
claim_expense <- 1500
expense_loading <- 0.10
```

The mortality curve below uses a Gompertz-Makeham-style force of mortality.
The parameters are deliberately illustrative and are not calibrated to an
official insured-life table.

```{r}
#| label: mortality-functions
make_mortality <- function(ages, mortality_factor = 1) {
  makeham_term <- 0.00035
  age_scale <- 0.000018
  growth <- 1.085

  force_of_mortality <- makeham_term + age_scale * growth^(ages - 30)
  qx <- 1 - exp(-force_of_mortality * mortality_factor)

  pmin(qx, 0.95)
}

survival_start <- function(qx) {
  if (length(qx) == 1) {
    return(1)
  }

  c(1, cumprod(1 - qx[-length(qx)]))
}

ages <- issue_age + seq_len(term) - 1
qx <- make_mortality(ages)

mortality_table <- data.frame(
  policy_year = seq_len(term),
  attained_age = ages,
  qx = round(qx, 6),
  deaths_per_1000 = round(1000 * qx, 3)
)

knitr::kable(head(mortality_table, 8))
```

### Premium basis

```{r}
#| label: actuarial-present-value-functions
term_assurance_apv <- function(qx, interest_rate) {
  years <- seq_along(qx) - 1
  discount_factor <- 1 / (1 + interest_rate)
  survival <- survival_start(qx)

  sum(discount_factor^(years + 1) * survival * qx)
}

annuity_due_apv <- function(qx, interest_rate) {
  years <- seq_along(qx) - 1
  discount_factor <- 1 / (1 + interest_rate)
  survival <- survival_start(qx)

  sum(discount_factor^years * survival)
}

benefit_apv <- sum_assured * term_assurance_apv(qx, annual_interest)
annuity_apv <- annuity_due_apv(qx, annual_interest)
net_level_premium <- benefit_apv / annuity_apv

expense_apv <- annual_expense * annuity_apv +
  claim_expense * term_assurance_apv(qx, annual_interest)

gross_level_premium <- (benefit_apv + expense_apv) /
  annuity_apv * (1 + expense_loading)

premium_summary <- data.frame(
  metric = c(
    "Issue age",
    "Term",
    "Annual effective discount rate",
    "Sum assured",
    "Net annual level premium",
    "Illustrative gross annual premium"
  ),
  value = c(
    issue_age,
    paste(term, "years"),
    format_percent(annual_interest),
    format_currency(sum_assured),
    format_currency(net_level_premium),
    format_currency(gross_level_premium)
  )
)

knitr::kable(premium_summary)
```

The gross premium is included only to make the projected portfolio cash-flow
view more realistic. It includes simple per-policy expenses and an illustrative
loading. The reserve calculation that follows is based on the net premium
formula, which isolates the mortality and discounting mechanics.

### Portfolio cash-flow projection

```{r}
#| label: cash-flow-projection
survival <- survival_start(qx)
inforce_start <- initial_policies * survival
expected_deaths <- inforce_start * qx

cash_flows <- data.frame(
  policy_year = seq_len(term),
  attained_age = ages,
  inforce_start = inforce_start,
  qx = qx,
  expected_deaths = expected_deaths,
  expected_premiums = inforce_start * gross_level_premium,
  expected_claims = expected_deaths * sum_assured,
  expected_maintenance_expenses = inforce_start * annual_expense,
  expected_claim_expenses = expected_deaths * claim_expense
)

cash_flows$expected_outgo <- with(
  cash_flows,
  expected_claims + expected_maintenance_expenses + expected_claim_expenses
)

cash_flows$expected_net_cash_flow <- with(
  cash_flows,
  expected_premiums - expected_outgo
)

v <- 1 / (1 + annual_interest)
cash_flows$pv_expected_premiums <- cash_flows$expected_premiums *
  v^(cash_flows$policy_year - 1)
cash_flows$pv_expected_outgo <- cash_flows$expected_outgo *
  v^cash_flows$policy_year
cash_flows$pv_margin <- cash_flows$pv_expected_premiums -
  cash_flows$pv_expected_outgo

display_cash_flows <- cash_flows
currency_columns <- c(
  "expected_premiums",
  "expected_claims",
  "expected_outgo",
  "expected_net_cash_flow"
)

display_cash_flows[currency_columns] <-
  lapply(display_cash_flows[currency_columns], format_currency)

display_cash_flows$inforce_start <- round(display_cash_flows$inforce_start, 1)
display_cash_flows$qx <- round(display_cash_flows$qx, 6)
display_cash_flows$expected_deaths <- round(display_cash_flows$expected_deaths, 3)

knitr::kable(head(display_cash_flows, 8))
```

```{r}
#| label: plot-cash-flows
#| fig-cap: "Expected annual cash flows for the illustrative term assurance portfolio."
plot(
  cash_flows$policy_year,
  cash_flows$expected_premiums / 1e6,
  type = "l",
  lwd = 2,
  col = "#1b998b",
  xlab = "Policy year",
  ylab = "Expected amount (R millions)",
  ylim = range(c(
    cash_flows$expected_premiums,
    cash_flows$expected_claims,
    cash_flows$expected_outgo
  )) / 1e6,
  main = "Projected portfolio cash flows"
)
lines(cash_flows$policy_year, cash_flows$expected_claims / 1e6,
      lwd = 2, col = "#2b4865")
lines(cash_flows$policy_year, cash_flows$expected_outgo / 1e6,
      lwd = 2, col = "#d49c2f")
legend(
  "topleft",
  legend = c("Premiums", "Claims", "Claims + expenses"),
  col = c("#1b998b", "#2b4865", "#d49c2f"),
  lwd = 2,
  bty = "n"
)
```

### Prospective reserve development

```{r}
#| label: reserve-functions
prospective_reserve <- function(
    issue_age,
    term,
    duration,
    interest_rate,
    sum_assured,
    annual_premium,
    mortality_factor = 1) {

  remaining_term <- term - duration

  if (remaining_term <= 0) {
    return(0)
  }

  attained_ages <- issue_age + duration + seq_len(remaining_term) - 1
  future_qx <- make_mortality(attained_ages, mortality_factor)

  future_benefit_apv <- sum_assured *
    term_assurance_apv(future_qx, interest_rate)
  future_premium_apv <- annual_premium *
    annuity_due_apv(future_qx, interest_rate)

  future_benefit_apv - future_premium_apv
}

durations <- 0:term

reserve_per_policy <- vapply(
  durations,
  prospective_reserve,
  numeric(1),
  issue_age = issue_age,
  term = term,
  interest_rate = annual_interest,
  sum_assured = sum_assured,
  annual_premium = net_level_premium
)

reserve_table <- data.frame(
  duration = durations,
  attained_age = issue_age + durations,
  reserve_per_policy = reserve_per_policy
)

reserve_display <- reserve_table
reserve_display$reserve_per_policy <-
  format_currency(reserve_display$reserve_per_policy)

knitr::kable(reserve_display[seq(1, nrow(reserve_display), by = 4), ])
```

```{r}
#| label: plot-reserve-trajectory
#| fig-cap: "Prospective net reserve per in-force policy by duration."
plot(
  reserve_table$duration,
  reserve_table$reserve_per_policy,
  type = "l",
  lwd = 3,
  col = "#2b4865",
  xlab = "Duration",
  ylab = "Reserve per policy",
  main = "Prospective reserve trajectory",
  yaxt = "n"
)
axis(2, at = pretty(reserve_table$reserve_per_policy),
     labels = format_currency(pretty(reserve_table$reserve_per_policy)))
abline(h = 0, lty = 2, col = "grey70")
points(
  reserve_table$duration,
  reserve_table$reserve_per_policy,
  pch = 19,
  col = "#1b998b"
)
```

The reserve starts close to zero because the net level premium is calibrated on
the same basis used for valuation. It then reflects the interaction between
ageing mortality, remaining premium income, and the shortening outstanding
term.

### Sensitivity testing

In valuation work, the base result is rarely the full story. A useful model
should help the reviewer understand how the liability behaves when the basis
changes.

The table below keeps the original net premium fixed and recalculates the
issue-date reserve under alternative mortality and discount-rate assumptions.
Positive values indicate that the valuation basis is stronger than the pricing
basis, before allowing for any explicit risk adjustment or margin.

```{r}
#| label: sensitivity-testing
valuation_margin <- function(discount_rate, mortality_factor) {
  stressed_qx <- make_mortality(ages, mortality_factor)

  stressed_benefit_apv <- sum_assured *
    term_assurance_apv(stressed_qx, discount_rate)
  stressed_premium_apv <- net_level_premium *
    annuity_due_apv(stressed_qx, discount_rate)

  stressed_benefit_apv - stressed_premium_apv
}

sensitivity <- expand.grid(
  discount_rate = c(0.065, 0.085, 0.105),
  mortality_factor = c(0.90, 1.00, 1.10, 1.25)
)

sensitivity$issue_date_margin <- mapply(
  valuation_margin,
  sensitivity$discount_rate,
  sensitivity$mortality_factor
)

sensitivity_display <- sensitivity
sensitivity_display$discount_rate <-
  format_percent(sensitivity_display$discount_rate)
sensitivity_display$mortality_factor <-
  paste0(sensitivity_display$mortality_factor, "x")
sensitivity_display$issue_date_margin <-
  format_currency(sensitivity_display$issue_date_margin)

knitr::kable(sensitivity_display)
```

## Actuarial Interpretation

The model highlights several valuation principles that matter in life insurance
practice:

- The premium basis and valuation basis must be separated clearly. A level
  premium calculated on one basis can create a liability or asset when assessed
  on another basis.
- Mortality, discounting, and timing conventions drive the result even in a
  simplified model. Small changes in assumptions can produce visible movement
  in the reserve profile.
- Portfolio-level cash-flow projections help connect policy-level actuarial
  formulae to management reporting, business planning, and capital modelling.
- A clean implementation should expose assumptions directly rather than burying
  them in opaque spreadsheet logic.

For an IFRS 17 or SAM context, this type of model is best viewed as a transparent
analytical layer. It can support reasonableness checks, sensitivity analysis,
and communication with finance or risk teams, even when the official valuation
engine is more complex.

## Model Limitations

This article intentionally excludes several features that would be required in
a production valuation model:

- lapses, surrenders, and paid-up policy behaviour;
- acquisition expenses, commission structures, tax, and reinsurance cash flows;
- stochastic economic scenarios and yield-curve construction;
- policyholder options and guarantees;
- risk adjustment, contractual service margin, and onerous group testing under
  IFRS 17; and
- governance controls such as assumption versioning, reconciliations, and model
  change logs.

Those omissions are deliberate. The purpose of the article is to present a
compact, auditable R implementation of the core mechanics before layering on
product complexity.
